Is it time for an NBA expansion?

### 2024-04-20 paper/paper.qmd
New Data and Model

library(tidyverse)
library(knitr)
library(modelsummary)
library(tinytex)
library(lubridate)
library(bibtex)
```

@@

NBA1980 <- read_csv("../data/raw_data/NBA1980.csv")
NBA1981 <- read_csv("../data/raw_data/NBA1981.csv")
NBA1982 <- read_csv("../data/raw_data/NBA1982.csv")
NBA1983 <- read_csv("../data/raw_data/NBA1983.csv")
NBA1984 <- read_csv("../data/raw_data/NBA1984.csv")
NBA1985 <- read_csv("../data/raw_data/NBA1985.csv")
NBA1986 <- read_csv("../data/raw_data/NBA1986.csv")
NBA1987 <- read_csv("../data/raw_data/NBA1987.csv")
NBA1988 <- read_csv("../data/raw_data/NBA1988.csv")
NBA1989 <- read_csv("../data/raw_data/NBA1989.csv")
NBA1990 <- read_csv("../data/raw_data/NBA1990.csv")
NBA1991 <- read_csv("../data/raw_data/NBA1991.csv")
NBA1992 <- read_csv("../data/raw_data/NBA1992.csv")
NBA1993 <- read_csv("../data/raw_data/NBA1993.csv")
NBA1994 <- read_csv("../data/raw_data/NBA1994.csv")
NBA1995 <- read_csv("../data/raw_data/NBA1995.csv")
NBA1996 <- read_csv("../data/raw_data/NBA1996.csv")
NBA1997 <- read_csv("../data/raw_data/NBA1997.csv")
NBA1998 <- read_csv("../data/raw_data/NBA1998.csv")
NBA1999 <- read_csv("../data/raw_data/NBA1999.csv")
NBA2000 <- read_csv("../data/raw_data/NBA2000.csv")
NBA2001 <- read_csv("../data/raw_data/NBA2001.csv")
NBA2002 <- read_csv("../data/raw_data/NBA2002.csv")
NBA2003 <- read_csv("../data/raw_data/NBA2003.csv")
NBA2004 <- read_csv("../data/raw_data/NBA2004.csv")
NBA2005 <- read_csv("../data/raw_data/NBA2005.csv")
NBA2006 <- read_csv("../data/raw_data/NBA2006.csv")
NBA2007 <- read_csv("../data/raw_data/NBA2007.csv")
NBA2008 <- read_csv("../data/raw_data/NBA2008.csv")
NBA2009 <- read_csv("../data/raw_data/NBA2009.csv")
NBA2010 <- read_csv("../data/raw_data/NBA2010.csv")
NBA2011 <- read_csv("../data/raw_data/NBA2011.csv")
NBA2012 <- read_csv("../data/raw_data/NBA2012.csv")
NBA2013 <- read_csv("../data/raw_data/NBA2013.csv")
NBA2014 <- read_csv("../data/raw_data/NBA2014.csv")
NBA2015 <- read_csv("../data/raw_data/NBA2015.csv")
NBA2016 <- read_csv("../data/raw_data/NBA2016.csv")
NBA2017 <- read_csv("../data/raw_data/NBA2017.csv")
NBA2018 <- read_csv("../data/raw_data/NBA2018.csv")
NBA2019 <- read_csv("../data/raw_data/NBA2019.csv")
NBA2020 <- read_csv("../data/raw_data/NBA2020.csv")
NBA2021 <- read_csv("../data/raw_data/NBA2021.csv")
NBA2022 <- read_csv("../data/raw_data/NBA2022.csv")
NBA2023 <- read_csv("../data/raw_data/NBA2023.csv")

@@

# Loop through the years 1980 to 2023
for(year in 1980:2023) {
  # Read the CSV file for the year
  file_path <- sprintf("../data/raw_data/NBA%d.csv", year)
  yearly_data <- read_csv(file_path)

@@

Define $y_i$ as the average number of points per game scored by a team through out the NBA season. Then $\alpha$ is the average assists per game, $\rho$ the average rebounds per game, $\beta$ is blocks per game, $\psi$ is steals per game and lastly, $\tau$ is turnovers per game.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \rho_i + \beta_i + \xi_i + \tau_i \\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\rho &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\psi &\sim \mbox{Normal}(0,2.5) \\
\tau &\sim \mbox{Normal}(0,2.5) \\
\sigma &\sim \mbox{Exponential}(1) \\
\end{align}

```{r}
#| warning: false
#| message: false

# Fit a multiple linear regression model
nba_model <- lm(PTS ~ Year * Num_Teams + AST + TRB + STL + BLK + TOV, data = nba_data)
nba_model_2 <- lm(PTS ~ Year + Num_Teams + AST + TRB + STL + BLK + TOV, data = nba_data)
model_ny <- lm(PTS ~ Num_Teams + AST + TRB + STL + BLK + TOV, data = nba_data)

plot(model_ny)
# Summary of the linear model
msummary(
  list(
    "times Year" = nba_model,
    "Without Year" = model_ny,
    "plus year" = nba_model_2
  ),
  fmt = 2
  )

summary(model_ny)
summary(nba_model_2)
```

@@

# Create histogram for count of records per year for scores of 60 or more from 1980 onwards
ggplot(data_1980_60pts_plus, aes(x = Year)) +
  geom_histogram(stat = "count", bins = length(unique(data_1980_60pts_plus$Year)), fill = "red", color = "black") +
  labs(title = "Count of Records per Year for Scores of 60+ PTS from 1980 Onwards",
       x = "Year Modified",
       y = "Count of Records") +
  theme_minimal()
```

```{r}

@@

>>>>

library(tidyverse)
library(knitr)
library(modelsummary)
library(tinytex)
library(lubridate)
```

@@

NBA1980 <- read_csv("../data/raw_data/NBA1980.csv")
NBA1981 <- read_csv("../data/raw_data/NBA1981.csv")
NBA1982 <- read_csv("../data/raw_data/NBA1982.csv")
NBA1983 <- read_csv("../data/raw_data/NBA1983.csv")
NBA1984 <- read_csv("../data/raw_data/NBA1984.csv")
NBA1985 <- read_csv("../data/raw_data/NBA1985.csv")
NBA1986 <- read_csv("../data/raw_data/NBA1986.csv")
NBA1987 <- read_csv("../data/raw_data/NBA1987.csv")
NBA1988 <- read_csv("../data/raw_data/NBA1988.csv")
NBA1989 <- read_csv("../data/raw_data/NBA1989.csv")
NBA1990 <- read_csv("../data/raw_data/NBA1990.csv")
NBA1991 <- read_csv("../data/raw_data/NBA1991.csv")
NBA1992 <- read_csv("../data/raw_data/NBA1992.csv")
NBA1993 <- read_csv("../data/raw_data/NBA1993.csv")
NBA1994 <- read_csv("../data/raw_data/NBA1994.csv")
NBA1995 <- read_csv("../data/raw_data/NBA1995.csv")
NBA1996 <- read_csv("../data/raw_data/NBA1996.csv")
NBA1997 <- read_csv("../data/raw_data/NBA1997.csv")
NBA1998 <- read_csv("../data/raw_data/NBA1998.csv")
NBA1999 <- read_csv("../data/raw_data/NBA1999.csv")
NBA2000 <- read_csv("../data/raw_data/NBA2000.csv")
NBA2001 <- read_csv("../data/raw_data/NBA2001.csv")
NBA2002 <- read_csv("../data/raw_data/NBA2002.csv")
NBA2003 <- read_csv("../data/raw_data/NBA2003.csv")
NBA2004 <- read_csv("../data/raw_data/NBA2004.csv")
NBA2005 <- read_csv("../data/raw_data/NBA2005.csv")
NBA2006 <- read_csv("../data/raw_data/NBA2006.csv")
NBA2007 <- read_csv("../data/raw_data/NBA2007.csv")
NBA2008 <- read_csv("../data/raw_data/NBA2008.csv")
NBA2009 <- read_csv("../data/raw_data/NBA2009.csv")
NBA2010 <- read_csv("../data/raw_data/NBA2010.csv")
NBA2011 <- read_csv("../data/raw_data/NBA2011.csv")
NBA2012 <- read_csv("../data/raw_data/NBA2012.csv")
NBA2013 <- read_csv("../data/raw_data/NBA2013.csv")
NBA2014 <- read_csv("../data/raw_data/NBA2014.csv")
NBA2015 <- read_csv("../data/raw_data/NBA2015.csv")
NBA2016 <- read_csv("../data/raw_data/NBA2016.csv")
NBA2017 <- read_csv("../data/raw_data/NBA2017.csv")
NBA2018 <- read_csv("../data/raw_data/NBA2018.csv")
NBA2019 <- read_csv("../data/raw_data/NBA2019.csv")
NBA2020 <- read_csv("../data/raw_data/NBA2020.csv")
NBA2021 <- read_csv("../data/raw_data/NBA2021.csv")
NBA2022 <- read_csv("../data/raw_data/NBA2022.csv")
NBA2023 <- read_csv("../data/raw_data/NBA2023.csv")
NBA2024 <- read_csv("../data/raw_data/NBA2024.csv")

@@

# Loop through the years 1980 to 2023
for(year in 1980:2024) {
  # Read the CSV file for the year
  file_path <- sprintf("../data/raw_data/NBA%d.csv", year)
  yearly_data <- read_csv(file_path)

@@

Define $y_i$ as the average number of points per game scored by a team through out the NBA season. Then $\alpha$ is the average assists per game, $\rho$ the average rebounds per game, $\beta$ is blocks per game, $\psi$ is steals per game and lastly, $\tau$ is turnovers per game, $\iota$ is the year, and $\eta$ is the number of teams.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \rho_i + \beta_i + \xi_i + \tau_i + \iota_i + \eta_i \\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\rho &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\psi &\sim \mbox{Normal}(0,2.5) \\
\tau &\sim \mbox{Normal}(0,2.5) \\
\iota &\sim \mbox{Normal}(0,2.5) \\
\eta &\sim \mbox{Normal}(0,2.5) \\
\sigma &\sim \mbox{Exponential}(1) \\
\end{align}

```{r}
#| warning: false
#| message: false
#| label: fig-predict
#| fig-cap: PPG Predecition Through the 2027 Season

# Fit a multiple linear regression model
nba_model <- lm(PTS ~ Year + AST + TRB + STL + BLK + TOV + Num_Teams, data = nba_data)

create_future_data <- function(additional_teams) {
  latest_data <- tail(nba_data, 1)
  adjustment_factor <- 1 - 0.03 * additional_teams  # Each new team reduces stats by 3%
  data.frame(
    Year = c(2024, 2025, 2026),
    Num_Teams = latest_data$Num_Teams + additional_teams,
    AST = rep(latest_data$AST * adjustment_factor, 3),
    TRB = rep(latest_data$TRB * adjustment_factor, 3),
    STL = rep(latest_data$STL * adjustment_factor, 3),
    BLK = rep(latest_data$BLK * adjustment_factor, 3),
    TOV = rep(latest_data$TOV * adjustment_factor, 3)
  )
}

# Create data frames for each scenario
data_no_new_teams = create_future_data(0)
data_one_new_team = create_future_data(1)
data_two_new_teams = create_future_data(2)

predict_points <- function(data, scenario_name) {
  predictions <- predict(nba_model, newdata = data)
  return(data.frame(Teams = scenario_name, Year = c(2025, 2026, 2027), Predicted_PTS = predictions))
}

# Generate predictions for each scenario
predictions_no_new_teams <- predict_points(data_no_new_teams, "30 Teams")
predictions_one_new_team <- predict_points(data_one_new_team, "31 Teams")
predictions_two_new_teams <- predict_points(data_two_new_teams, "32 Teams")

# Combine predictions into a single data frame
all_predictions <- bind_rows(predictions_no_new_teams, predictions_one_new_team, predictions_two_new_teams)

# Reshape the data frame
wide_predictions <- all_predictions %>%
  pivot_wider(names_from = Year, values_from = Predicted_PTS, values_fill = list(Predicted_PTS = NA))

# Print the reshaped data frame
wide_predictions

@@

# Create histogram for count of records per year for scores of 60 or more from 1980 onwards
ggplot(data_1980_60pts_plus, aes(x = Year)) +
  geom_histogram(stat = "count", bins = length(unique(data_1980_60pts_plus$Year)), fill = "red", color = "black") +
  labs(title = "Count of Records per Year for Scores of 70+ PTS from 1980 Onwards",
       x = "Year Modified",
       y = "Count of Records") +
  theme_minimal()
```

@@

msummary(
  list(
    "Points Model" = nba_model
  ),
  fmt = 2
  )
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| 
summary(nba_model)
```
